\prefacesection{Abstract}

Building a large library of mathematical knowledge is a complex and labour-intensive task.
By examining current libraries of mathematics, we see that the human effort put in building them is not entirely directed towards tasks that need human creativity. 
Instead, a non-trivial amount of work is spent on providing definitions that could have been mechanically derived.

%Instead, a non-trivial amount is directed towards redefining basic mathematical knowledge. 
%Instead of focusing all the human effort on formalizing more knowledge, we find that some basic mathematical concepts are being redefined again and again based on changing design decisions. 

In this work, we propose a generative approach to library building, so definitions that can be automatically derived are computed by a meta-program. We focus our attention on libraries of algebraic structures, like monoids, groups, and rings. 
These structures are highly inter-related and their commonalities have been well-studied in universal algebra. We use theory presentation combinators to build a library of algebraic structures. Definitions from universal algebra and programming languages meta-theory are used to derive library definitions of constructions, like homomorphism and term languages, from algebraic theory presentations. The result is an interpreter that, given $227$ theory expressions, builds a library of over $5000$ definitions. This library is, then, exported to Agda and Lean. 

%We use meta-programs to generate  

%where definitions that are normally provided by library developers are automatically generated by a meta-program. We focus our attention to algebra libraries. Algebraic theories are highly structured and their commonalities has been well-studied in universal algebra. We use theory presentation combinators to capture the structure of algebraic theories. Universal algebra definitions are used to derive definitions of constructions, like homomorphism and term languages, from algebraic theory presentations. The result is an interpreter that, given $227$ theory expressions, builds a library of over $5000$ definitions.


