\prefacesection{Acknowledgements}
\vspace{-1cm}
\begin{center} 
\textit{Praise and gratitude be to Allah, the most gracious and the most beneficent.}
%with the blessings whom the good deeds are fulfilled.
\end{center}
\noindent I would like to express my sincere thanks to my supervisors Dr.~Jacques Carette and Dr.~William Farmer for their continuous support to my learning journey.  Your expertise and feedback were invaluable in shaping this research direction and throughout my studies. 
I learned from you a lot about how to do research and communicate it. 
I am very thankful to Dr.~Wolfram Kahl and Dr.~Ridha Khedri for being on my supervisory committee. Your feedback, over the years, on this work is very appreciated. Many thanks to my external examiner, Dr. Makarius Wenzel, for the useful feedback and the interesting discussions. 

I am grateful to Dr.~Michael Kohlhase, Dr.~Florian Rabe, and members of the KWARC team for inviting me to their research group at FAU Erlangen. The collaboration between our teams has been a great source of inspiration for this work. 

I would like to express of my heartfelt appreciation to Dr. Andrea Kohlhase for her support and for being a great role model. 

Without the support of my family and friends, I wouldn't have been able to pursue this degree. Thanks for all the love and support. Thank you for always believing in me and encouraging me to explore different areas. 
I am most grateful to Ahmad, my husband, best friend and mentor. Thank you for always being there for me. I am so proud of the team we make. 
